Determining Causes of Death with ChatGPT: A Case Study of Verbal Autopsy Data in Sierra Leone from 2019-2022

### 2024-01-04 manuscript/manuscript.tex
Add to bg

\title[Determining Causes of Death with ChatGPT: A Case Study of Deaths from Verbal Autopsy in Sierra Leone from 2019-2022]{Determining Causes of Death with ChatGPT: A Case Study of Deaths from Verbal Autopsy in Sierra Leone from 2019-2022}

@@

Reliable counts and diagnosis of deaths are crucial for public health planning and policy making --- guiding scalable interventions that save lives and reduce premature deaths worldwide \cite{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, most low-income countries have reported to have either no data on deaths or have registered less than 50\% of deaths in their country, where an even fewer 8\% of these registered deaths have a Cause of Death (COD) recorded \cite{worldhealthorganizationSCOREHealthData2021}. To fill in this gap, an alternative method known as Verbal Autopsy (VA) is used to collect data on deaths and determine their likely causes at scale, outside of traditional healthcare facilities \cite{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014}.

VA involves two major components: survey and COD assignment \cite{worldhealthorganizationVerbalAutopsyStandards2023}. In the survey component, trained lay surveyors interview those familiar with the deceased (e.g. living spouse, family, relatives, friends) to gather information using standardized questionnaires and open narratives. The gathered information consists of a standardized questionnaire, and an open narrative. In the COD assignment component, physicians evaluate information gathered from lay surveyors to assign probable CODs. Although the survey component has been an effective alternative to collect mortality data at scale, the COD assignment component has been criticized to be expensive and difficult to reproduce due to reliance on physician assigned CODs \cite{mccormickProbabilisticCauseofDeathAssignment2016,morrisFactorsAssociatedPhysician2010}. Recently, computer algorithms have been studied to automatically assign CODs as an alternative to physicians, with performances close to physicians at the population level, but poor performances at the individual level ((CITE)). Many of these computer algorithms have utilized the questionnaire portion of the surveyed data, but often omit data from the open narrative.

>>>>

\title[Determining Causes of Death with ChatGPT: A Case Study of Verbal Autopsy Data in Sierra Leone from 2019-2022]{Determining Causes of Death with ChatGPT: A Case Study of Verbal Autopsy Data in Sierra Leone from 2019-2022}

@@

In 2019, 41 million people die prematurely from noncommunicable diseases every year, accounting for 74\% of all deaths globally \cite{worldhealthorganizationNonCommunicableDiseases2019}. Most of these deaths are preventable and treatable, but require adequate resources, interventions, and policies, such as access to primary care, health insurance, healthier diets, and advanced screening procedures \cite{benzigerGlobalBurdenDisease2016}. Reliable counts and diagnosis of deaths are crucial for public health planning and policy making --- guiding scalable interventions that save lives and reduce premature deaths worldwide \cite{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, most low-income countries do not have data on deaths or registered less than half of the deaths in their country, with an even fewer 8\% of these registered deaths having a Cause of Death (COD) recorded \cite{worldhealthorganizationSCOREHealthData2021}. To fill this gap in death registrations, an alternative method known as Verbal Autopsy (VA) is used to collect data on deaths and determine their likely causes at scale \cite{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014}, outside of traditional healthcare facilities where over half of the deaths occur at home \cite{adairWhoDiesWhere2021}.

VA involves two major components: survey and COD assignment \cite{worldhealthorganizationVerbalAutopsyStandards2023}. In the survey component, trained lay surveyors interview those familiar with the deceased (e.g. living spouse, children, family, friends) to gather information using standardized questionnaires and open narratives. In the COD assignment component, physicians evaluate information available from the questionnaires and open narratives to assign probable CODs. Although the survey component has been an effective alternative to collect mortality data at scale, the COD assignment component has been criticized to be expensive and difficult to reproduce due to reliance on physician assigned CODs \cite{mccormickProbabilisticCauseofDeathAssignment2016,morrisFactorsAssociatedPhysician2010}. Recently, computer algorithms have been studied to automatically assign CODs, as an alternative to physician assignment, with performances close to physicians at the population level, but poor performances at the individual level \cite{}. Many of these computer algorithms have utilized the questionnaire portion of the surveyed data, but often omit data from the open narrative.
